---
title: "Project2"
author: "YuMei, Anthony & Nitin"
date: "March 31, 2018"
output: html_document
---

Github repository
https://github.com/tbitxyz/SMU-DDS-Machine-Churning-Project-2
---

-- Executive Summary --  
---
The analysis uses the casestudy2Data provided to evaluate the ralationship between employee attrition and factors such as; Education, EnvironmentSatisfaction, JobInvolvement, JobSatisfaction , PerformanceRating , RelationshipSatisfaction , WorkLifeBalance e.t.c . The response variable is "attrition" which has two levels. Therefore, Random Forrest method was used to classify the data and identify the factors that are strongly associated with employee attrition. 

Before apply the Random Forest analysis, we employed scatter plot and bar chart to explore the data itself, making sure the data is fit for random forest analysis.

Two type of random forest R packages are utilized for analysis. The result are not identical, but pretty close. Both showed that Overtime is strongly associated with Attrition, folowed by Age and length of services. 

-- Introduction --  
---
Any business is modeled for profit-making, and the management unit of any organization is tasked with ensuring that the goal is accomplished. It is therefore important for the human resource department to ensure that the personnel functions optimally.  One method used in ensuring that the personnel delivers as expected is monitoring employee performance from time to time and developing employee performance monitoring scale.  Sometimes, it might be necessary for an organization to take disciplainary action if the goals of the business are not met.  


```{r Attrition analysis, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

# Data Analysis predicting Employee Attrition
dir()
casestudy <- read.csv("E:\\University\\SMU\\Doing the Data Science\\Project2\\CaseStudy2data.csv", header = TRUE)
head(casestudy)
View(casestudy)

casestudy$attri <- ifelse(casestudy$Attrition == "Yes", 1,0)

casestudy$educ <- factor(casestudy$Education, levels = c(1,2,3,4,5), labels = c("Below College","College","Bachelor","Master","Doctor"))

casestudy$environsat <- factor(casestudy$EnvironmentSatisfaction, levels = c(1,2,3,4), labels = c("Low","Medium","High","very High"))

casestudy$jobinv  <- factor(casestudy$JobInvolvement, levels = c(1,2,3,4), labels = c("Low","Medium","High","very High"))

casestudy$jobsat <- factor(casestudy$JobSatisfaction, levels = c(1,2,3,4), labels = c("Low","Medium","High","very High"))

casestudy$perfrate  <- factor(casestudy$PerformanceRating, levels = c(1,2,3,4), labels = c("Low","Good","Excellent","Outstanding"))

casestudy$relationsat <- factor(casestudy$RelationshipSatisfaction, levels = c(1,2,3,4), labels = c("Low","Medium","High","very High"))

casestudy$workbal <- factor(casestudy$WorkLifeBalance, levels = c(1,2,3,4), labels = c("Bad","Good","Better","Best"))


# descriptive Analysis of the data

table(casestudy$Attrition)  # Number of employee who experienced attrition from the sample

# Attrition by Education Level
table(casestudy$Attrition, casestudy$educ)

t1 <- table(casestudy$Attrition, casestudy$educ)
t1
barplot(t1, legend = row.names(t1), beside = TRUE)

t2 <- table(casestudy$Attrition, casestudy$environsat)  # Environment satisfaction and Attrition. 
t2
barplot(t2,legend = row.names(t2), beside = TRUE, main = "Environment satisfation by Attrition")

# Performance rate and Attrition

t3 <- table(casestudy$Attrition, casestudy$perfrate)
t3

barplot(t3,legend = row.names(t3), beside = TRUE, main = "perfomance rate by Attrition")

# Job Satisfaction and Attrition

t4 <- table(casestudy$Attrition, casestudy$jobsat)
t4
barplot(t4,legend = row.names(t4), beside = TRUE, main = "Job Satisfaction by Attrition")

# Job involvement by Attrition

t5 <- table(casestudy$Attrition, casestudy$jobinv)
t5
barplot(t5,legend = row.names(t5), beside = TRUE, main = "Job Involvement by Attrition")

# relationship satisfaction and Attrition

t6 <- table(casestudy$Attrition, casestudy$relationsat)
t6
barplot(t6,legend = row.names(t6), beside = TRUE, main = "Relationsip satisfaction by Attrition")


# EXAMINING THE EFFECT OF EDUCATION STATUS
# Education status by Performance rating

t7 <- table(casestudy$educ, casestudy$perfrate)
t7
barplot(t7,legend = row.names(t7), beside = TRUE, main = "Performance rating by Education level")


#  Job Satisfaction by Education level

t8 <- table(casestudy$educ, casestudy$jobsat)
t8
barplot(t8,legend = row.names(t8), beside = TRUE, main = "Job Satisfaction by Education level")
```


```{r project2, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
# read in dataset
turnover<-read.csv("E:\\University\\SMU\\Doing the Data Science\\Project2\\CaseStudy2data.csv", header=TRUE,na.strings=c(""))
# find out missing value
sapply(turnover,function(x) sum(is.na(x)))
# no missing values, but first column name looks odd.
# find Column name "Age"
require(dplyr)
turnover1 = rename(turnover, Age = 'ï..Age')
head(turnover1)
# checking all variables before apply analytics
str(turnover1)
# remove NAs if any
turnover2 <- na.omit(turnover1)

## extract factor columns and drop variable with single level
which(sapply(turnover2, function(x) { length(unique(x)) }) == 1)
turnover3 <- turnover2[,-c(27,22,9)] 

## count levels
sapply(turnover3, nlevels)
str(turnover3)

# drop varibles that does not make sense for Attrition
turnover4<-turnover3[-match("EmployeeNumber",names(turnover3))]
head(turnover4)

#remove DailyRate, HourlyRate and MonthlyRate, these variables are meaningful for company, not employee
turnover51<- turnover4[, -grep("Rate$", colnames(turnover4))]

# are any varibles correlated?
turnover52 <- turnover51[sapply(turnover51,is.numeric)]
pairs(turnover52[,1:4], pch = 19)
# don't look like anything here
pairs(turnover52[,5:8], pch = 19)
# job level and monthly income have a strong relationship
pairs(turnover52[,9:12], pch = 19)
# performance rating and salary hike looks has a tight relationship
pairs(turnover52[,13:16], pch = 19)
# nothing here
pairs(turnover52[,17:20], pch = 19)
# we see visible correlation between 4 varibles that starts with Years. 
str(turnover52)
#finding few of them on scatter plot, attemptt to further verify with cor function
cor(turnover52$Age,turnover52)
cor(turnover52$PerformanceRating,turnover52)
cor(turnover52$YearsAtCompany,turnover52)

# combine varibles with strong coorlation
AgeIncomeLevel<-turnover52$Age+turnover52$JobLevel*turnover52$MonthlyIncome
RatingHike<-turnover52$PerformanceRating*turnover52$PercentSalaryHike
Years<-turnover52$YearsAtCompany+turnover52$TotalWorkingYears+turnover52$YearsInCurrentRole+turnover52$YearsSinceLastPromotion+turnover52$YearsWithCurrManager
# drop original varibles
turnover53 = subset(turnover51, select = -c(Age,JobLevel,MonthlyIncome,PerformanceRating,PercentSalaryHike,YearsAtCompany,TotalWorkingYears,YearsInCurrentRole,YearsSinceLastPromotion,YearsWithCurrManager) )
# add combined varibles
turnover5<-cbind(turnover53, AgeIncomeLevel,RatingHike,Years)

# use cforest to find outt what varibble are important to Attrition prediction 
library(party)
cf1 <- cforest(Attrition ~ . , data= turnover5, control=cforest_unbiased(mtry=5,ntree=50)) # fit the random forest
cf1
importance<-varimp(cf1) # get variable importance, based on mean decrease in accuracy
#  plot importance
importance1<-sort(importance, decreasing = TRUE)
library(ggplot2)
importance2 <- data.frame(x=names(importance1), y=importance1)
importance3<-arrange(importance2, desc(y))

ggplot(importance3, aes(x,y, fill=x)) + geom_bar(stat="identity")+ coord_flip()+
  geom_text(aes(y=max(y), label=y), color="black")+
  scale_fill_discrete(guide="none")+
  theme_bw()+theme(panel.grid=element_blank())

# cforest importance analysis shows "over time" is the number one contributing factor for Attrition. Followed by Years and AgeIncomeLevel. 

AgeIncomeLevel<-turnover52$Age+turnover52$JobLevel*turnover52$MonthlyIncome
RatingHike<-turnover52$PerformanceRating*turnover52$PercentSalaryHike
Years<-turnover52$YearsAtCompany+turnover52$TotalWorkingYears+turnover52$YearsInCurrentRole+turnover52$YearsSinceLastPromotion+turnover52$YearsWithCurrManager

# try random forest to verify the result
library(randomForest)

model.rf <- randomForest(Attrition~., turnover5, ntree=100, proximity=TRUE, importance=TRUE)
meande<-importance(model.rf, type=1) #preferred

meande4<-c(meande)
names(meande4)<-c("BusinessTravel", "Department", "DistanceFromHome","Education", "EducationField", "EnvironmentSatisfaction", "Gender", "JobInvolvement", "JobRole", "JobSatisfaction", "MaritalStatus", "NumCompaniesWorked", "OverTime", "RelationshipSatisfaction", "StockOptionLevel", "TrainingTimesLastYear", "WorkLifeBalance", "AgeIncomeLevel", "RatingHike", "Years")

meande5 <- data.frame(x=names(meande4), y=meande4)
meande6<-arrange(meande5, desc(y))

ggplot((meande6), aes(x,y, fill=x)) + geom_bar(stat="identity")+ coord_flip()+
  geom_text(aes(y=max(y), label=y), color="black")+
  scale_fill_discrete(guide="none")+
  theme_bw()+theme(panel.grid=element_blank())

# This clearly agrees with cforest package analysis result, that is "over time" is the dominant factor for Attrition. There are few secondary factors include "Years" which is a aggregate of several length  of service varibles; "AgeIncomeLevel" which is another aggregate varible for Age, Job Level and Monthly Income.  


```

